Adaptive evolution with aneuploidy and mutation

### 2022-03-27 ms/ms.tex
bit more TODOS finished
Update ms.tex

Transitions between the genotypes occur as follows~(\Cref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate. We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\; and from \anmt\ to \eumt\ with probability $\delta$, the aneuploidy rate. That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
The fitness values of the four genotypes are given by \Cref{table:fitness}.

%%% Table: fitness
\begin{table}[h]
\centering
\caption{\textbf{Fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^*$ & $2n^*$ \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)(1+s)+b$ & $1+s$               
\end{tabular}
\label{table:fitness}
\caption*{
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$0 \le c \le 1$ is the fitness cost of aneuploidy;
and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%

The initial population has $N$ cells with genotype \euwt.
The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
    \begin{equation} \label{eq:selection-single} 
      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
    \end{equation}
where the fitness values $w_i$ are given in \Cref{table:fitness} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
The effect of mutation and aneuploidy on genotype frequencies is given by
    \begin{equation} \label{eq:mutation-aneuploidy-single}
    \begin{aligned}
      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \mu) f^s_{2n+1}  \;,\\
      &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
    \end{aligned}
    \end{equation}
Finally, random genetic drift is modeled using a multinomial distribution~\citep{Otto2007},
    \begin{equation} \label{eq:drift-single}
      \vec{f'} \sim \frac{1}{N} \cdot \mathit{Mult}(N,\ \vec{f^m}) \;,
    \end{equation}
where $\vec{f^m}=(f^m_{2n}\ ,\ f^m_{2n+1}\ ,\ f^m_{2n+1^*}\ ,\ f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N,\ \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.

@@

For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), we disregard the increased expression in chromosome III and the growth advantage measured in generation 450, and focus on the growth advantage measured in later generations, presumably due to a beneficial mutation. 
Therefore, the likelihood is approximated by
\begin{equation}\begin{aligned}
\label{eq:heatstress-noaneuploidy-likelihood}
\likelihood_{!}(\theta) = &\ 
	1 - 
	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\}\big) - \\
&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\}\big) + \\
&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\}\big)
\;.
\end{aligned}\end{equation}

@@

\paragraph{Model comparison.} The main criterion that we use for model selection is WAIC, the widely applicable information criterion ~\citep{gelman2013bayesian},
\begin{equation} \label{eq:WAIC}
\begin{aligned}
\mathit{WAIC}(\theta) &\ =\ 
-2\log\mathbb{E}[\likelihood(\theta)]\ +\ 2\mathbb{V}[\log\likelihood(\theta)]
\end{aligned}
\end{equation}
where $\theta$ is a parameter vector, and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution, which in practice are approximated using 50,000 samples from the posterior KDE.
WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.

We also plot posterior predictions: for each model we execute $10,000$ simulations using the MAP parameter estimates and plot the distributions of time to fixation of \eumt, one of key properties of the model likelihood. These plots visualize the fit of each model to the data. 
Also, for similar models we plot the marginal and joint posterior distributions of the parameters; if these are similar, we consider the models are interchangeable. We validate this by comparing HDI (highest density interval) of posterior distributions.

\paragraph{Prior distributions.} We used informative prior distributions for $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs. 3C, 4A, and S2]{Yona2012}.
We used \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}. Briefly, \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
From these predictions, relative fitness values can be computed. Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix.

We sampled 10,000 samples to use as a prior distribution~(\Cref{fig:growth-curves}).
We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$~(\Cref{fig:growth-curves}-D, assuming $w_{2n}=1$). In lieu of a more suitable prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$. 
To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results. 
We also examined an extended informative prior that uses additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$~(\Cref{fig:growth-curves}L). The same prior was used for $w_{2n*}/w_{2n+1*}$.

As a control, we tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$, or (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$. In these cases the inference algorithm failed to converge.

@@

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of model parameters~(\Cref{fig:posterior}). 
We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets.
%% estimates generated by MAP-and-hdi.ipynb
The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$, agrees with previous estimates. % TODO refs (9.7*10-5 in Zhu 2014, so it's lower). YR: we need more than one ref; also what is the units of Zho 2014?
The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
The estimated fitness values are $w_{2n+1}=1.022\ [1.021-1.023]$,
$w_{2n+1^*}=1.025\ [1.024-1.026]$,
$w_{2n^*}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$. 
Thus, we can infer that the benefit and cost of trisomy was $b=2.5\%$ and $c=0.3\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{table:fitness}).

@@

% FIG posterior alt-prior
%% generated with alt-prior.ipynb
%% TODO IK - update the figure to look like Figure 2
\begin{figure}[p]
  \centering
\includegraphics[width=0.9\textwidth]{../figures/posterior-alt.pdf}
  \caption{
  \textbf{Posterior distribution of parameters inferred with the extended prior distribution.}
On the diagonal, the inferred posterior distribution of each model parameter. 
Below the diagonal, the inferred joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively). Red markers and orange lines for the joint MAP estimate (which may differ from the marginal MAP, as the marginal distribution integrates over all other parameters).
} 
  \label{fig:posterior-alt}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%

\paragraph{Model checking and comparison.}
The model fits the data well: in simulations using the MAP parameter estimates, \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\Cref{fig:2n*-fixation}B). 
Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\Cref{fig:ppc-plot}).

Moreover, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}). %% generated with MAP-and-hdi.ipynb
The estimated mutation rate without aneuploidy is $\mu=7.979\cdot10^{-9}\ [7.903\cdot 10^{-9}-8.135\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40. 
The fitness of the mutant is also much lower at $w_{2n^*}=1.013\ [1.012-1.013]$.
This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations. Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).

@@

Sensitivity analysis shows that changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\Cref{fig:sensitivity}).
Furthermore, we fitted models with a mutation rate fixed at $\mu$=$10^{-5}$, $10^{-6}$ and $10^{-7}$. We inferred similar parameters estimates for the model with $\mu=10^{-6}$ compared to the model with a free $\mu$ parameter. Models with $\mu$=$10^{-5}$ and $\mu$=$10^{-7}$ inferred different parameters estimates~(\Cref{fig:mu}).
WAIC was lower when $\mu$ is fixed~(\Cref{table:WAIC}), but this is not surprising, as WAIC attempt to balance between model fit and model complexity, where the latter takes into account the number of model parameters.

% generated with alt-prior.ipynb
Finally, we estimated the parameters under an extended informative prior (see \emph{Prior distributions}), with different priors for $w_{2n+1^*}$ and $w_{2n*}$. The estimated mutation rate was much lower compared to the basic prior, with $\mu=2.379\cdot10^{-9}\ [2.267\cdot10^{-9}-2.458\cdot10^{-9}]$. Other parameter estimates are: $\delta=2.636\cdot10^{-3}\ [2.245\cdot10^{-3}-2.919\cdot10^{-3}]$,
$w_{2n+1}=1.022\ [1.021-1.023]$,
$w_{2n+1^*}=1.032\ [1.03-1.035]$,
$w_{2n^*}=1.033\ [1.031-1.036]$. 
The WAIC of the model is much lower: -57 compared to 295~(\Cref{table:WAIC}).

% Fig model with and without aneuploidy
%% generated with with-aneuploidy_vs_no_aneuploidy.ipynb
\begin{figure}[p]
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/fixation-plot-a.pdf}      
      \label{fig:fit}
  \end{subfigure}
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/fixation-plot-b.pdf}      
      \label{fig:fit-cumulative}
  \end{subfigure}
  \caption{
    \textbf{Model fit with and without aneuploidy.}
    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green).
In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450. Thus, the blue distribution is a better fit compared to the green, and the yellow histogram has a very poor fit.
The last bin contains all the simulations with time equal or greater than 3,000.
    \textbf{(B)}
	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively, and corresponding to the blue and green bars in panel A. The MAP likelihood (\Cref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
     }
     % TODO should we remove this figure altoghther? change it? because we removed no-aneuploidy model...? should we merge this with the next figure?
  \label{fig:2n*-fixation}
\end{figure}

\begin{figure}[p]
  \centering
\includegraphics[width=0.8\textwidth]{../figures/dynamics.pdf}
  \caption{
  \textbf{Posterior genotype frequency dynamics for the model.} 
  The posterior prediction for the frequencies of the four genotypes over time. Each of the 20 curves is the average of 10,000 simulations of the model using parameters drawn from the posterior distribution.
  }
  \label{fig:ppc-plot}
\end{figure}

% Fig tau plots
%% generated by diff-tau-plots.ipynb
\begin{figure}[p]
  \centering
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-a.pdf}      
  \end{subfigure}
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-b.pdf}      
  \end{subfigure}
   \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-c.pdf}      
  \end{subfigure}
   \caption{
    \textbf{Genotype fixations for models with different $\tau$.} We estimated the parameters for different models, each assuming a different value of $\tau$, the fold-increase in mutation rate in aneuploid cells. We then generated 10,000 simulations using the MAP estimate of each model and evaluated the fraction of simulations in which the genotype \euwt\ (\textbf{A}), \anwt\ (\textbf{B}), and \eumt\ (\textbf{C}) is fixed (y-axis) at each generation (x-axis). Note that \anmt\ did not fix. We can see that $\tau=100$ can be distinguished if time to loss of \euwt\ is known (panel~A) or if time to fixation or loss of \anwt\ is known (panel~B). It is harder to distinguish between $1\leq \tau \leq10$.
  }
  \label{fig:tau-plots}
\end{figure}

@@

>>>>

Transitions between the genotypes occur as follows~(\Cref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate. We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\; and from \anmt\ to \eumt\ with probability $\delta$, the aneuploidy rate. That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
The fitness values of the four genotypes are given by \Cref{table:single-locus}.

%%% Table: single-locus model
\begin{table}[h]
\centering
\caption{\textbf{Single-locus model fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^*$ & $2n^*$ \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)(1+s)+b$ & $1+s$               
\end{tabular}
\label{table:single-locus}
\caption*{
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$0 \le c \le 1$ is the fitness cost of aneuploidy;
and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%

The initial population has $N$ cells with genotype \euwt.
The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
    \begin{equation} \label{eq:selection-single} 
      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
    \end{equation}
where the fitness values $w_i$ are given in \Cref{table:single-locus} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
The effect of mutation and aneuploidy on genotype frequencies is given by
    \begin{equation} \label{eq:mutation-aneuploidy-single}
    \begin{aligned}
      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \mu) f^s_{2n+1}  \;,\\
      &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
    \end{aligned}
    \end{equation}
Finally, random genetic drift is modeled using a multinomial distribution~\citep{Otto2007},
    \begin{equation} \label{eq:drift-single}
      \vec{f'} \sim \frac{1}{N} \cdot \mathit{Mult}(N,\ \vec{f^m}) \;,
    \end{equation}
where $\vec{f^m}=(f^m_{2n}\ ,\ f^m_{2n+1}\ ,\ f^m_{2n+1^*}\ ,\ f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N,\ \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.

@@

% YR: this whole section doesn't really make sense. We know there was aneuploidy. If we assume delta=0 that means that we think this aneuploidy was not relevant to the adaptation process. But then why do we take generations 1700 and 2300 to be the time for 2n* fixation? These times were not observed for fixation of a mutant but rather for elimination of aneuploidy. If we disregard aneuploidy, why do we still consider these times?
%For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), 
%the likelihood is similarly approximated by
%\begin{equation}\begin{aligned}
%\label{eq:heatstress-noaneuploidy-likelihood}
%\likelihood_{!}(\theta) = &\ 
%	1 - 
%	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\}\big) - \\
%&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\}\big) + \\
%&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\}\big)
%\;.
%\end{aligned}\end{equation}

@@

\paragraph{Model comparison.} We perform model selection using WAIC, the widely applicable information criterion ~\citep{gelman2013bayesian},
\begin{equation} \label{eq:WAIC}
\begin{aligned}
\mathit{WAIC}(\theta) &\ =\ 
-2\log\mathbb{E}[\likelihood(\theta)]\ +\ 2\mathbb{V}[\log\likelihood(\theta)]
\end{aligned}
\end{equation}
where $\theta$ is a parameter vector, and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution, which in practice are approximated using 50,000 samples from the posterior KDE.
WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
% TODO IK Describe posterior plots comparison method - Fig 4 & S3

\paragraph{Prior distributions.} We used informative prior distributions for $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs. 3C, 4A, and S2]{Yona2012}.
We used \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}. Briefly, \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
From these predictions, relative fitness values can be computed. Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix. We sampled 10,000 samples to use as a prior distribution~(\Cref{fig:growth-curves}).
We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\Cref{fig:growth-curves}-D). In lieu of a more suitable prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$. 
To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results.

Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
First, we tried to use additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$ ~(\Cref{fig:growth-curves}L). The same prior was used for $w_{2n*}/w_{2n+1*}$. This prior resulted in WAIC 71.91, compared to 67.87 with the above informative prior. % generated with fixed-mutation-rate-and-WAICs.ipynb TODO IK update the WAIC with 1000 reps run
We also tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$, or (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$. In both cases the inference algorithm failed to converge.

@@

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of model parameters~(\Cref{fig:posterior}). 
We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets.
%% estimates generated by MAP-and-hdi.ipynb
The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$, agrees with previous estimates. % TODO refs (9.7*10-5 in Zhu 2014, so it's lower). YR: we need more than one ref; also what is the units of Zho 2014?
The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
The estimated fitness values are $w_{2n+1}=1.022\ [1.021-1.023]$,
$w_{2n+1^*}=1.025\ [1.024-1.026]$,
$w_{2n^*}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$. 
Thus, we can infer that the benefit and cost of trisomy was $b=2.5\%$ and $c=0.3\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{table:single-locus}).

@@

\paragraph{Model checking and comparison.}
The model fits the data well: in simulations using the MAP parameter estimates, \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\Cref{fig:2n*-fixation}B). % TODO IK what happened to the sentence about the fixation on anwt? green lines ub fig4?
Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\Cref{fig:ppc-plot}).

% commented out, see remark above when describing likelihood of delta=0.
%Moreover, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}). %% generated with MAP-and-hdi.ipynb
%The estimated mutation rate without aneuploidy is $\mu=7.979\cdot10^{-9}\ [7.903\cdot 10^{-9}-8.135\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40. 
%The fitness of the mutant is also much lower at $w_{2n^*}=1.013\ [1.012-1.013]$.
%This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations. Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).

@@

Sensitivity analysis shows that changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\Cref{fig:sensitivity}).
Furthermore, we fitted models with a mutation rate fixed at $\mu$=$10^{-5}$, $10^{-6}$ and $10^{-7}$. We inferred similar parameters estimates compared to the model with a free $\mu$ parameter. % TODO IK where is this shown? how do you quantify it?
WAIC was lower when $\mu$ is fixed~(\Cref{table:WAIC}), but this is not surprising, as WAIC attempt to balance between model fit and model complexity, where the latter takes into account the number of model parameters.

% Fig single-locus model with and without aneuploidy
%% generated with with-aneuploidy_vs_no_aneuploidy.ipynb
\begin{figure}[p]
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/fixation-plot-a.pdf}      
      \label{fig:fit}
  \end{subfigure}
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/fixation-plot-b.pdf}      
      \label{fig:fit-cumulative}
  \end{subfigure}
  \caption{
    \textbf{Single-locus model fit with and without aneuploidy.}
    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green).
In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450. Thus, the blue distribution is a better fit compared to the green, and the yellow histogram has a very poor fit.
The last bin contains all the simulations with time equal or greater than 3,000.
    \textbf{(B)}
	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively, and corresponding to the blue and green bars in panel A. The MAP likelihood (\Cref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
     }
  \label{fig:2n*-fixation}
\end{figure}

\begin{figure}[p]
  \centering
\includegraphics[width=0.8\textwidth]{../figures/dynamics.pdf}
  \caption{
  \textbf{Posterior genotype frequency dynamics for the single-locus model.} 
  The posterior prediction for the frequencies of the four genotypes over time. Each of the 20 curves is the average of 10,000 simulations of the single-locus model using parameters drawn from the posterior distribution.
  }
  \label{fig:ppc-plot}
\end{figure}

% Fig tau plots
%% generated by diff-tau-plots.ipynb
\begin{figure}[p]
  \centering
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-a.pdf}      
  \end{subfigure}
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-b.pdf}      
  \end{subfigure}
   \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-c.pdf}      
  \end{subfigure}
   \caption{
    \textbf{Genotype fixations for models with different $\tau$.} We estimated the parameters for different models, each assuming a different value of $\tau$, the fold-increase in mutation rate in aneuploid cells. We then generated 10,000 simulations using the MAP estimate of each model and evaluated the fraction of simulations in which the genotype \euwt\ (\textbf{A}), \anwt\ (\textbf{B}), and \eumt\ (\textbf{C}) is fixed (y-axis) at each generation (x-axis). Note that \anmt\ did not fix. We can see that $\tau=100$ can be distinguished if time to loss of \euwt\ is known (panel~A) or if time to fixation or loss of \anwt\ is known (panel~B). It is harder to distinguish between $1\leq \tau \leq10$.
  }
  \label{fig:tau-plots}
  % TODO IK axis labels like in figure 3, also somehow the graphics looks different from previous figures
  % fixed means 100% ?
\end{figure}

@@

\paragraph{Approximation.}
We aim to approximate the adaptation time (i.e., time to fixation of \eumt) and analyse how it is affected by the model parameters.
There are two stochastic events that can occur on the \euwt\ genotype background. The first is a mutation that produces a \eumt\ genotype.
The second is a mis-segregation event that produces a \anwt\ genotype.
The rate at which these events occur is significantly different, and we assume that if one event occurs (e.g., mutation), then the other event (e.g., aneuploidy) is unlikely to occur.
The expected waiting time until one of these two genotype appears and then leads to adaptation is therefore
\begin{equation}\label{eq:T_total}
T = p_a T_a+(1-p_a) T_m \;,
\end{equation}
where $p_a$ is the probability that aneuploidy is the first event to occur, 
$T_m$ is the expected time for appearance and fixation of \eumt\ given it appears before \anwt, and $T_a$ is the expected time for appearance of \anwt, followed by a mutation to \anmt, and loss of aneuploidy to \eumt, given that \anwt\ appears before \eumt.

We assume the waiting time $T_{i,j}$ for appearance of genotype $j$ in a population of genotype $i$ is geometrically distributed with rate $p_{i,j}$, such that
\begin{subequations}
\begin{align}
p_{2n, 2n^*} &= 1-\Big(1 - 2 \frac{w_{2n^*}-w_{2n}}{w_{2n}} \mu \Big)^N \;, \\
p_{2n, 2n+1} &= 1-\Big(1 - 2 \frac{w_{2n+1}-w_{2n}}{w_{2n}} \delta \Big)^N \; \\
p_{2n+1, 2n+1^*} &= 1-\Big(1 - 2 \frac{w_{2n+1^*}-w_{2n+1}}{w_{2n+1}} \mu \Big)^N \; \\
p_{2n+1^*, 2n^*} &=1-\Big(1 - 2 \frac{w_{2n^*}-w_{2n+1^*}}{w_{2n+1^*}} \delta \Big)^N \;.
\end{align}
\end{subequations}
However, if the mis-segragation supply is high enough, $\delta N > 1$, then \eumt\ may appear before \anmt\ is fixed (\Cref{fig:ppc-plot}), and the latter equation is
\begin{equation}
p_{2n+1^*, 2n^*} = 1-\Big(1-2 \frac{w_{2n^*}-w_{2n+1}}{w_{2n+1}} \delta \Big)^{zN} \;,
\end{equation}
where $z \ll 1$ is the frequency of \anmt\ when \eumt\ first appears (note that here we use $w_{2n+1}$ rather than $w_{2n+1^*}$).

We assume that if \anwt\ appears in a population of \euwt\ and does not go to extinction then \eumt\ does not appear, and vice versa. Thus, we need to find the minimum of the waiting times for appearance of these genotypes. The minimum of two geometric random variables with rates $p_{2n,2n^*}$ and $p_{2n,2n+1}$ is also a geometric random variable with rate $p_{min} = 1-(1-p_{2n,2n^*})(1-p_{2n,2n+1})$, which is the inverse of the expected waiting time until the first event occurs, $T_{min}=1/p_{min}$.
Given that one of these genotypes appeared, the probability that it was \anwt\ rather than \eumt\ is
$p_A = \frac{p_{2n,2n+1}}{p_{2n,2n^*}+p_{2n,2n+1}}$.

Given a mutation occurs first, the expected adaptation time via the mutation trajectory is 
\begin{equation}\label{eq:T_a} 
T_{M} = T_{min} + \tau_{2n, 2n^*} \;,
\end{equation}
where $\tau_{i, j}$ is the waiting time for fixation of genotype $j$ in a population of genotype $i$, given it starts with a single copy and does not go to extinction by drift.
Given that aneuploidy occurs first,
the expected adaptation time via the aneuploidy trajectory is approximated by
\begin{equation}\label{eq:T_a} 
T_a = T_{min} + T_{2n+1, 2n+1^*} + T_{2n+1^*,2n^*} + \tau_{2n, 2n+1} + \tau_{2n+1, 2n+1^*} + \tau_{2n+1^*, 2n^*} \;.
\end{equation}

We assume that the waiting time until a new genotype occurs, $T{i,j}$, is much longer than the waiting time for its fixation, $\tau_{i,j}$. Nevertheless, these fixation times must be included in a rigorous analysis. There are two ways to compute the fixation times.
A deterministic-process approach is given by the classical bi-allelic single-locus equation.
It approximates the expected time for genotype $j$ to reach frequency $x$ when starting from frequency $y$ in a population of genotype $i$ by
 % TODO MP ref for equation? Crow and Kimura? YR: yes
\begin{equation} \label{eq:tau_fix_x}
\tau_{i, j}(x,y) \approx \frac{\log{\Big(\frac{x(1-y)}{(1-x)y} \Big)}}{\log{\Big(\frac{w_j}{w_i}\Big)}} \;,
\end{equation} 
and we use $\tau_{i,j} = \tau_{i,j}(1-1/N, 1/N) \approx 2\log{(N)}/\log{(w_j/w_i)}$, except for $\tau_{i,2n^*} = \tau_{i,2n^*}(0.95, 1/N)$ because we stop our simulations when genotype \eumt\ reaches frequency of 0.95.
A stochastic-process approach for approximating $T_{i, j}$ is given by the diffusion-equation approximation \citep[eq.~17]{Kimura1969a}. 
% TODO MP give details here
